Is structure based drug design ready for selectivity optimization?

### 2018-12-19 manuscript.tex
first draft done

%
%
%  Discussion and Conclusions 
%
%
\section{Discussion and Conclusions}
We have demonstrated, using a simple numerical model, the impact that free energy calculations with even weakly correlated errors can have on speeding up the optimization of selectivity in small molecule kinase inhibitors. While the expected speed up is dependent on the per target error of the method ($\sigma_{target}$), the speedup is also highly dependent on the correlation of errors made for both targets. Unsurprisingly, free energy methods have greater impact as the threshold for selectivity optimization goes from 10x to 100x. While 100x selectivity optimization is difficult to achieve, the expected benefit from free energy calculations is also quite high, with 1 and 2 order of magnitude speedups possible. 
To quantify the correlation of errors in two example systems, we gathered experimental data for two congeneric ligand series with experimental data for CDK2 and ERK2, as well as CDK2 and CDK9. These datasets, which had crystal structures for both targets with the same ligand co-crystallized, exemplify the difficulty in predicting selectivity. The dynamic range of selectivity for both systems is incredibly narrow, with most of the perturbations not having a major impact on the overall selectivity achieved. Further, the data was reported with unrelaible experimental uncertainties, which makes quantifying the errors made by the free energy calculations difficult. This issue is common when considering selectivity, as many kinase-oriented high throughput screens are carried out at a single concentration and not highly quantitative.

>>>>

%
%
%  Discussion and Conclusions 
%
%

\section{Discussion and Conclusions}
There are a number of different metrics for quantifying the selectivity of a compound~\citep{Bosc:2017gs}, which look at selectivity from different views depending on the information trying to be conveyed. One of the earliest metrics was the standard selectivity score, which conveyed the number of inhibited kinase targets in a broad scale assay divided by the total number of kinases in the assay~\citep{Davis2011-dz}. The gini coefficient is a method that does not rely on any threshold, but is highly sensitive to experimental conditions because it is dependent on percent inhibition~\citep{Graczyk:2007bm}. Other metrics take a thermodynamic approach to kinase selectivity and are suitable for smaller panel screens~\citep{DuongLy:2016iha,Uitdehaag:2011ea}. Here, we propose a more granular, thermodynamic  view of selectivity that is easy to use free energy methods to calculate: the change in free energy of binding for a given ligand between two different targets ($\Delta \Delta G_{selectivity}$). $\Delta \Delta G_{selectivity}$ is a useful metric of selectivity in lead optimization once a single, or small panel, of off-targets have been identified and the goal is to use physical modeling to either improve or maintain selectivity within a lead series.

We have demonstrated, using a simple numerical model, the impact that free energy calculations with even weakly correlated forcefield errors can have on speeding up the optimization of selectivity in small molecule kinase inhibitors. While the expected speed up is dependent on the per target forcefield error of the method ($\sigma_{ff}$), the speedup is also highly dependent on the correlation of errors made for both targets. Unsurprisingly, free energy methods have greater impact as the threshold for selectivity optimization goes from 10x to 100x. While 100x selectivity optimization is difficult to achieve, the expected benefit from free energy calculations is also quite high, with 1 and 2 order of magnitude speedups possible.

To quantify the correlation of errors in two example systems, we gathered experimental data for two congeneric ligand series with experimental data for CDK2 and ERK2, as well as CDK2 and CDK9. These datasets, which had crystal structures for both targets with the same ligand co-crystallized, exemplify the difficulty in predicting selectivity. The dynamic range of selectivity for both systems is incredibly narrow, with most of the perturbations not having a major impact on the overall selectivity achieved. Further, the data was reported with unrelaible experimental uncertainties, which makes quantifying the errors made by the free energy calculations difficult. This issue is common when considering selectivity, as many kinase-oriented high throughput screens are carried out at a single concentration and not highly quantitative. Despite CDK2 and ERK2 being more distantly related than CDK2 and CDK9, the calculated correlation in the forcefield error suggests that fortuitous cancellation of errors may be applicable in a wider range of scenarios than closely related kinases within the same family.

We built a numerical model of the impact of statistical error in the context of different levels of forcefield error correlation, in order to better understand if there are situations where it is beneficial to expend more effort running longer calculations to minimize statistical error and get improved speedup in selectivity optimization. Our results suggest that unless the correlation is above 0.5 for the two targets of interest, there is not much benefit in running longer calculations. However, when the forcefield error is reduced by correlation, longer calculations can help realized large increases in speedup. Keeping a running quantification of $\rho$ for free energy calculations as compounds are made and the predictions can be tested will allow for decisions to be made about whether running longer calculations is worthwhile. It will also allow for an estimate of $\sigma_{selectivity}$, which is useful for estimating expected forcefield error for prospective predictions. Importantly, we expect that correlation will be protocol dependent and changes to the way the system is modeled are expected to change the observed correlation in the forcefield error.

This work demonstrates that correlation in the forcefield errors can allow free energy calculations to facilitate significant speedups in selectivity optimization for drug discovery projects. This is particularly important in kinase systems, where considering multiple targets is an important part of the process. The results suggest that free energy calculations can be particularly helpful in the design of kinase polypharmacological agents, especially in cases where there is high correlation in the forcefield errors between multiple targets.
